Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-09 Paper/paper.tex
working on s3

\newcommand{\ii}{\mathrm{i}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\deff}{D_\mathrm{eff}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

Eq.~\ref{eq:marginalize} to obtain the marginalized probability distribution
for the tidal deformability of the NS $p(\lambdans|d_n,K)$.
The most likely value of $\lambdans$ for this distribution is our 
{\it measured} value for the parameter, and the $90\%$ confidence intervals
associated with the distribution serve as statistical error-bars.

@@

% We use this
% distribution to put confidence bounds on our measurement of $\lambdans$.
% of this distribution to estimate our statistical and

@@

\red{Reorder:
We construct NSBH waveforms with LEA+. We inject them, and run MCMC on the injections
using BBH templates. We perform this analysis for [... parameters ...]. We always
use $\mns=1.35M_\odot$, but explore $\lambdans=\{500,1000,1500,2000\}$.
}

\prayush{
We first take a set of NSBH binary waveforms, which include the modeling of spin and
tidal effects.
For all of the resulting signals, we perform a fully Bayesian parameter estimation
aimed at measuring binary masses and spins from the signal. Our analysis
templates differ from the signal in only one sense: all tidal terms are set to zero
in the former, i.e. they correspond to BHBH mergers. Binary mass-ratio is sampled
uniformly over the range $q\in[2,5]$ taking $q=\{2,3,4,5\}$ as our test points. The
mass of the NS is fixed to $\mns=1.35M_\odot$, and therefore knowing $q$ gives $\mbh$.
The spin of the NS is fixed to zero. The spin of the BH is sampled uniformly over the
range $\chibh\in[-0.75, +0.75]$, choosing $\chibh=\{-0.5, 0, +0.5, +0.75\}$ as our
test points. Finally, we vary the deformability parameter uniformly over 
$\lambdans\in[0,2000]$, choosing $\lambdans=\{500, 1000, 1500, 2000\}$ plus an additional
$\lambdans=800$ case. Waveform are generated for all possible combinations of these
parameters, and rescaled to different strengths with $\rho=\{20, 30, 50, 70\}$.
}

>>>>

\newcommand{\ii}{\mathrm{i}}
\newcommand{\chibh}{\chi_\mathrm{BH}}
\newcommand{\chins}{\chi_\mathrm{NS}}
\newcommand{\mbh}{m_\mathrm{BH}}
\newcommand{\mns}{m_\mathrm{NS}}
\newcommand{\mchirp}{\mathcal{M}_c}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\deff}{D_\mathrm{eff}}
\newcommand{\arr}{\mathcal{r}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques}\label{s1:techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% #################
% THESE FIGURES ARE FOR THE NEXT SECTION
\begin{figure*}
\centering 
\includegraphics[width=1.8\columnwidth]{plots/SingleSystemEta_q4_0_mc2_25_chi0_50}
\caption{We show here the measured probability distribution for the
dimensionless mass ratio $\eta$ of a binary with
$q = \mbh/\mns = 5.4M_\odot/1.35M_\odot = 4$, $\chibh=+0.5$, and
$\lambdans=2000$. The templates used here {\it ignore} tidal effects.
These figures show that this approximation holds up to (SNR)
$\rho\simeq 30-50$.
% 
In each panel- the dashed red line marks the {\it measured} value of $\eta$,
which is given by the median of its probability distribution.
The {\it true} value is marked by the dashed green line. The difference
between the true and measured values is the systematic bias introduced
by ignoring tidal effects in templates. The darker shading shows
the recovered $90\%$ confidence interval for $\eta$. The width of this
region is a direct measure of our statistical uncertainty.
% 
Comparing systematic and statistical errors, we find that:
at $\rho=20$, $\eta$ measurement is dominated by statistical
errors; at $\rho=30$, the two become comparable; and 
for louder signals ($\rho\simeq50$), the systematic errors dominate.
}
\label{fig:SingleSystemEtaPDFvsSNR}
\end{figure*}
% #################
\begin{figure*}
\centering 
\includegraphics[width=1.9\columnwidth]{plots/TNMchirpCIWidths90_0_Lambda_SNR}\\
\includegraphics[width=1.9\columnwidth]{plots/TNEtaCIWidths90_0_Lambda_SNR}\\
\includegraphics[width=1.9\columnwidth]{plots/TNChiBHCIWidths90_0_Lambda_SNR}
\caption{
We show here the statistical uncertainty of our measurement of the 
dimensionless mass ratio $\eta$ (at $90\%$ confidence), over the NSBH
parameter space.
Individual panels show the same as a function of BH mass and spin.
Across each row, we see the effect of increasing signal strength (i.e.
SNR) with the NS's tidal deformability $\lambdans$ fixed. Down each
column, we see the effect of increasing $\lambdans$, at fixed SNR.
}
\label{fig:CIWidths90_Lambda_SNR}
\end{figure*}
% ###############################
% In this section we summarize various technical aspects of this paper. The model
% used to obtain tidal waveforms is described in Sec.~\ref{s2:waveforms}. The 
% process of inferring source parameters from a GW signal is summarized in 
% Sec.~\ref{s2:bayesian}. The process of generating and combining multiple events
% is instead detailed in the corresponding results section,
% Sec.~\ref{s2:astro_multiple}.

@@

Eq.~\ref{eq:marginalize} to obtain marginalized probability distributions
for the NS tidal deformability parameter: $p(\lambdans|d_n,K)$.
The median value of this distribution is our {\it measured} value for
$\lambdans$, and the $90\%$ confidence intervals associated with the distribution
serve as statistical error-bars.

@@

@@

% \red{Reorder:
% We construct NSBH waveforms with LEA+. We inject them, and run MCMC on the injections
% using BBH templates. We perform this analysis for [... parameters ...]. We always
% use $\mns=1.35M_\odot$, but explore $\lambdans=\{500,1000,1500,2000\}$.
% }

We construct NSBH waveforms using the LEA+ model. We inject these into zero
noise at different SNRs, and run MCMC on the injections using BBH 
templates (LEA+ with tidal terms set to zero). We perform this analysis for mass
ratios $q\in\{2,3,4,5\}$, i.e.
$\mbh\in\{2.7M_\odot, 4.05M_\odot, 5.4M_\odot, 6.75M_\odot\}$, and BH spins
$\chibh\in\{-0.5, 0, +0.5, +0.75\}$. We always use $\mns=1.35M_\odot$ and
$\chins=0$, but explore $\lambdans\in\{500, 800, 1000, 1500, 2000\}$.

% We simulate NSBH signal waveforms using the LEA+ model, inject them into zero
% noise, and perform fully Bayesian parameter estimation analyses on them to
% estimate their source masses and spins. We set tidal terms in search
% templates to zero, ensuring that they mimic BBH mergers. We perform this
% analysis for a range of systems: with mass ratios $q\in\{2,3,4,5\}$, BH spins
% $\chibh\in\{-0.5, 0, +0.5, +0.75\}$, and NS tidal deformability
% $\lambdans\in\{500, 800, 1000, 1500, 2000\}$. We fix NS mass to $1.35M_\odot$
% and its spin to zero (for signals). In addition, signals are simulated at
% different strengths, i.e. $\rho=\{20, 30, 50, 70\}$, in order to estimate the
% dependence of parameter measurement errors on the SNR.

As an illustration, in Fig.~\ref{fig:SingleSystemEtaPDFvsSNR} we show the
probability distribution for binary mass ratio $\eta$ recovered from a GW
injection with intrinsic parameters: $\mns=1.35M_\odot$, $\chins=0$,
$\mbh=5.4M_\odot$, $\chibh=+0.5$ and $\lambdans=2000$, and SNRs ranging
from $\rho=20$ (left panel) to $\rho=50$ (right panel). In each panel, the
{\it true} value of $\eta$ is marked by a dashed green line, and its measured
(median) value by a dashed red line. The effect of ignoring tidal corrections
in templates manifests as a systematic shift of the measured $\eta$ value away
from its true value. The associated {\it systematic} bias/error is given by the
difference between measured and true $\eta$ values. Darker shading of the
probability distribution marks the $90\%$ confidence interval for $\eta$, whose
width is a direct measure of the {\it statistical} measurement uncertainty/error.
% 
We clearly see that even when the signal is moderately loud, with $\rho=20$,
statistical errors dominate over systematics completely for $\eta$.
As we turn up the SNR
further, we find that the two error sources become comparable at $\rho\sim 30$,
and systematic errors dominate finally when $\rho\simeq 50$.

Next, to get a better understanding of how {\it precisely} we measure NSBH
binary parameters given BBH waveforms, we show the measured $90\%$ confidence
intervals for binary masses and spins in Fig.~\ref{fig:CIWidths90_Lambda_SNR}.
The three panels, corresponding to $\mchirp$ (top), $\eta$ (middle), and $\chibh$
(bottom), show the width of these confidence intervals as a function of BH
mass/spin (within each sub-panel), and NS properties, i.e. $\lambdans$
(downwards in each column).~\footnote{We restrict NS mass to $1.35M_\odot$ and
its spin to zero. Varying its tidal deformability $\lambdans$ does not change
the measurement uncertainties for non-tidal binary parameters, as is evident
from comparing the two rows in each panel of Fig.~\ref{fig:CIWidths90_Lambda_SNR}.}
% 
From the left-most column, we find that: (i) at $\rho=20, \mchirp$ is already
measured remarkably well - to a precision of $0.16\%$ of its true value, and
(ii) so is $\chibh$. (iii) Mass ratio $\eta$ is determined somewhat worse,
with $\gtrsim 25\%$ uncertainty. If the signal is even louder ($\rho\geq 30$),
all three measurements gain further precision, with $\eta$ errors shrinking down
to single-digit percents of its true value.
% 
However, we remind ourselves that there are physical effects that have not been
incorporated in our (BBH) search templates which we use here to decipher NSBH
signals. Therefore, we cannot know the {\it accuracy} of our parameter
measurements unless we know how large the systematic biases are due to an
incomplete template model. To quantify this, we find it useful to define $\arr$
as the ratio of the systematic biases induced by unmodeled tidal effects in
templates, to the statistical uncertainties associated with the measurement
itself. If $\arr\gtrsim 1$, including the effect of NS's
tidal distortion (during inspiral) and possible disruption (close to merger)
would be imperative to yield reliable estimates of binary masses and spins
from its measured GW signal.
